Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis

### 2015-04-07 paper/main.tex
respond to reviewer's comments, Edward's take 1

The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.  A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple species across multiple time points.  We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

@@

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.  Our model, while focusing on predators' food selection not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.  Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers the formal statistical justification of GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.

@@

Our model is based on two Poisson distributions with parameters describing the rates at which the predator consumes prey species and the rate at which the predator encounters prey species.  Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.  Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

This paper is organised as follows.  Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the expectation-maximisation (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.  In Section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.  Section~\ref{sec:data} provides a real data set, which investigates the eating preferences of wolf spiders (Araneae: Lycosidae), found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how those interested in assessing trophic interactions with gut-content analyses could apply our methods.  A brief discussion concludes the paper in Section~\ref{sec:discussion}.  Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that implements the methods discussed.

@@

To illustrate these methods, we analysed a data set that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).  Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.  Spiders were removed from the leaf litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.  In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.  For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.  On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.  The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.  Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.

@@

With point estimates of $c_{st}$ under the model $c_{st} = \lambda_{st}/\gamma_{st}$, we can test any number of linear contrasts.  For instance, the hypotheses $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$ state that wolf spiders equally prefer the orders Diptera and Collembola at each of the $18$ time points.  Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.

@@

All the methods mentioned here, whether they be others' methods or our model, are susceptible to a number of issues that must be handled as best as possible by the investigator.  The most obvious are sampling issues of prey and predator species of interest.  It is almost impossible to know how well one's data represents the habitat and the availability of the prey of interest.  Variables not recorded when collecting data could have influenced the relative abundance of species that were caught.  And yet this is just the beginning.  Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.  We therefore, in this paper, focus solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.

The earliest attempts to estimate predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  focused on producing one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.  Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  We feel these indices missed the point; an index is only as good as it is understood to estimate and model our world of observable data.

@@

The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.  When appropriately fit to data our methods offer greater statistical power than any index or distribution-free Monte Carlo method could.  We were able to justify our method with robust simulations that provide insight into the statistical properties of our model.  Moreover, we offer the means to estimate parameters of interest even when the predator's gut contents are not fully observed.

Throughout this paper we assume the observed data were generated from Poisson distributions.  This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.  Of course other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.  For instance, consider simply replacing the Poisson distribution with a distribution with more than one parameter.  Since, in the case of non-count data, only $0$s and $1$s are observed, an underlying distribution with more than one parameter would be over-identified.  The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about underyling processes that are being modelled.

Need words about users of these methods.

>>>>

The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.  A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple prey species across multiple time points.  We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

@@

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.  Our model, while focusing on predators' food selection, not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.  Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.

@@

Our model is based on two Poisson distributions with parameters describing the rates at which the predator consumes prey species and the rate at which the predator encounters prey species.  Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.  When exact counts are unobserved, we instead rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

This paper is organised as follows.  Section~\ref{sec:methods} describes our statistical model, for both fully observed count data and for the non-observed count data for which we use the expectation-maximisation (EM) algorithm to estimate parameters of interest, and the statistical tests used to make statements about the parameters of interest.  In Section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.  Section~\ref{sec:data} provides a real data set, which investigates the eating preferences of wolf spiders (Araneae: Lycosidae), found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how those interested in assessing trophic interactions with gut-content analyses could apply our methods.  A brief discussion concludes the paper in Section~\ref{sec:discussion}.  Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that implements the methods discussed.

@@

To illustrate these methods, we analysed a data set collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).  Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.  Spiders were removed from the leaf litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.  In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.  For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.  On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.  The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.  Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.

@@

With point estimates of $c_{st}$ under the model $c_{st} = \lambda_{st}/\gamma_{st}$, we can test any number of linear contrasts.  For instance, the hypotheses $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$ state that wolf spiders equally prefer the orders Diptera and Collembola at each of the $18$ time points.  Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.  {\color{red} What does this tell us/someone? Just a quick sentence.}

@@

All the methods mentioned here, whether they be others' methods or our model, are susceptible to a number of issues that must be handled as best as possible by the investigator.  The most obvious are sampling issues of prey and predator species of interest.  It is almost impossible to know how well one's data represents the habitat and the availability of the prey of interest.  Variables not recorded when collecting data could have influenced the relative abundance of species that were caught.  And yet this is just the beginning.  Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.  We therefore, in this paper, focused solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.

The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  produced one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.  Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  We feel these indices missed the point; an index is only as good as it is understood to estimate and model our world of observable data.

@@

The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.  When appropriately fit to data our methods, via a likelihood ratio test, offer greater statistical power than any index or distribution-free Monte Carlo methods do.  Robust simulations provided insight into the statistical properties of our model.  Moreover, we can estimate parameters of interest even when the predator's gut contents are not fully observed.

Throughout this paper we assume the observed data were generated from Poisson distributions.  This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.  Of course, other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.  For instance, consider simply replacing the Poisson distribution with a distribution that has more than one parameter.  Since, in the case of non-count data only $0$s and $1$s are observed, a distribution with more than one parameter would be over-identified.  The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about the underyling processes that are being modelled.

{\color{red} So what does our model offer ecologists? Just a quick paragraph.}
